### 2014-08-07 Studies/2012_Chicago/Report/Chicago_Fire.tex
Chicago: Minor edits throughout

\chapter*{\centering Abstract}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on November 2, 2012, within a 2 \sfrac{1}{2} story, single family structure in Chicago, IL, that resulted in the death of the captain of a fire department. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at a height of 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.

@@

On November 2, 2012, a fire in a 2~\sfrac{1}{2} story wood frame residential structure claimed the life of a captain of the Chicago Fire Department. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{fire_sum} provides a summary of the fire incident, Section~\ref{model} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{results} presents the simulation results, and Section~\ref{discussion} discusses the simulation results as they relate to firefighter safety and effectiveness.

@@

Based on the information collected, several engineering approximations were made. The actual ignition time is unknown.

The following subsections will describe the inputs that were used to construct the simulation, including any approximations or assumptions that were made.

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2} story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~\ref{fig:simp_geom}. This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber covered by 12.7~mm (0.5~in) gypsum board~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~\ref{fig:simp_geom}), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~\ref{fig:simp_geom}) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~\ref{fig:simp_geom}). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~\ref{fig:simp_geom}.

@@

\begin{lstlisting}
&SPEC ID = 'WOOD', FORMULA = 'CH1.7O0.74N0.002' /

@@

\begin{lstlisting}
&SURF ID='FIRE', HRRPUA=2250., COLOR='ORANGE', TAU_Q=10 /
&VENT XB=-13.5,-13,1,5,6.16,6.16, SURF_ID='FIRE' / 
&VENT XB=-14.5,-14,1,5,6.16,6.16, SURF_ID='FIRE' / 
&VENT XB=-16.5,-15.5,1,2.5,3.14,3.14, SURF_ID='FIRE', DEVC_ID='SECOND FIRE'/ 
&DEVC XYZ = -17.2,2,0 ID = 'SECOND FIRE', QUANTITY = 'TIME', SETPOINT = 40, INITIAL_STATE = .FALSE. /
\end{lstlisting}
Note that the coordinates here are unique for the vents are specific to the input files used for these simulations. The key point is that they are 2-D planes located at a specific area.

@@

\chapter{Summary}
NIST's Fire Dynamics Simulator was used to provide insight into the fire dynamics within a 2 \sfrac{1}{2} story single family structure in Chicago, IL, that resulted in the death the captain of a fire department. The fuel, fire size, and fire growth rate that were used in the FDS simulations were estimated by taking into account all of the available information including the NIOSH post-incident report~\cite{NIOSH:Bowyer}, post-incident pictures, videos, and relevant literature. This resulted in a specified source fire of 12.3~MW. However, based on the limited ventilation conditions in the attic and enclosed porch, the FDS simulation results indicated a maximum HRR of approximately 9~MW as a result of the limited supply of oxygen.

>>>>

\chapter*{\centering Abstract}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on November 2, 2012, within a 2~\sfrac{1}{2}~story, single family structure in Chicago, IL, that resulted in the death of the captain of a fire department. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at a height of 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.

@@

On November 2, 2012, a fire in a 2~\sfrac{1}{2}~story wood frame residential structure claimed the life of a captain of the Chicago Fire Department. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{fire_sum} provides a summary of the fire incident, Section~\ref{model} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{results} presents the simulation results, and Section~\ref{discussion} discusses the simulation results as they relate to firefighter safety and effectiveness.

@@

The analysis of the simulation results is focused on the interior conditions on the second floor before and after the interior door failure. Based on the goals of the analysis and the information collected, several engineering approximations were made. For example, the fire exhibited signs of under-ventilated conditions when firefighters arrived~\cite{NIOSH:Bowyer}; however, the exact ignition time, fire growth rate, and heat release rate (HRR) of the fire were not known. The ignition time and growth rate were estimated to reduce the amount of required calculation time while adhering to documented observations. The HRR of the fire was estimated using documentation of the potential fuel load within the structure. Although there might be differences between the actual values and those used in the simulation, the model input parameters represent our best estimate of the fire scenario based on the information that was available. The following subsections describe the inputs that were used to develop the simulation, including any approximations or assumptions that were made.

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2}~story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~\ref{fig:simp_geom}. This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber covered by 12.7~mm (0.5~in) gypsum board~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~\ref{fig:simp_geom}), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~\ref{fig:simp_geom}) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~\ref{fig:simp_geom}). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~\ref{fig:simp_geom}.

@@

\begin{lstlisting}
&SPEC ID='WOOD', FORMULA='CH1.7O0.74N0.002' /

@@

\begin{lstlisting}
&SURF ID='FIRE', HRRPUA=2250., COLOR='ORANGE', TAU_Q=10 /
&VENT XB=-13.5,-13,1,5,6.16,6.16, SURF_ID='FIRE' / 
&VENT XB=-14.5,-14,1,5,6.16,6.16, SURF_ID='FIRE' / 
&VENT XB=-16.5,-15.5,1,2.5,3.14,3.14, SURF_ID='FIRE', DEVC_ID='SECOND FIRE'/ 
&DEVC XYZ = -17.2,2,0, ID='SECOND FIRE', QUANTITY='TIME', SETPOINT=40, INITIAL_STATE=.FALSE. /
\end{lstlisting}
Note that the coordinates here are unique for the vents are specific to the input files used for these simulations. The key point is that they are 2-D planes located at a specific area.

@@

\chapter{Summary}
NIST's Fire Dynamics Simulator was used to provide insight into the fire dynamics within a 2~\sfrac{1}{2}~story single family structure in Chicago, IL, that resulted in the death the captain of a fire department. The fuel, fire size, and fire growth rate that were used in the FDS simulations were estimated by taking into account all of the available information including the NIOSH post-incident report~\cite{NIOSH:Bowyer}, post-incident pictures, videos, and relevant literature. This resulted in a specified source fire of 12.3~MW. However, based on the limited ventilation conditions in the attic and enclosed porch, the FDS simulation results indicated a maximum HRR of approximately 9~MW as a result of the limited supply of oxygen.
